Modelling the emergence of whisker barrels

### 2020-05-28 paper/briefpaper/brief.tex
Thursday work on figures and paper

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers.  However, the
assumption of pre-arranged center-points is difficult to resolve with the
observation that axons arrive in the cortical plate as an undifferentiated
bundle, \emph{prior} to barreloid formation \citep{agmon_organized_1993}.

@@

\begin{figure}
\begin{fullwidth}
%\includegraphics[width=\linewidth]{./MainFig.png}
\includegraphics[width=\linewidth]{./MainFig_superlowres.png}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
  by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
  measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
  perfect Voronoi tesselation. Right shows the initial distribution of axon
  branching density ($a$) for one thalamocortical projection, and two
  molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction
  $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
  projections by the lengths of green and blue arrows respectively, assuming
  that similar fields aligned to the posterior-anterior and medial-lateral
  axes in the ventroposterior medial nucleus of the thalamus are sampled at
  the locations of putative barreloid centers (reconstructed from
  \citealp{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
  $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
  projection for which the connection density is maximal, black lines
  delineate boundaries, and overlaid contours show $c>0.5$ (see Movie
  S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
  simulation approaching that obtained from barrels in \textbf{A} (dotted
  line); black squares measure the correspondence between the real and
  simulated barrel shapes, $\eta$ (the product of the sum of squared
  differences between real and simulated centers and the sum of differences in
  area; units mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern
  in a Fgf8 misexpression experiment
  (c.f. \citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
  $\rho_1$ at the join ($\epsilon=80$). All scale bars 1mm.}
\label{fig:main}
\end{fullwidth}
\end{figure}

@@

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. As such, two orthogonal linear thalamic
gradients were defined, from which 41 pairs of $\gamma$ values were sampled,
at the coordinates of 41 barreloid centers recreated from
\cite{haidarliu_size_2001} (Fig.\,1B).

A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a(\mb{x},0)$ and $c(\mb{x},0)$, a clear Voronoi-like tesselation emerged
(Fig.\,1C; see Movie S1). A reduction in the Honda delta metric (see
\citealp{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged
within $\approx 3000$ iterations (Fig.\,1D). A measure of the difference
in shape between real and simulated barrels revealed a strong correspondence
(see Fig.\,1D).

\begin{figure}
\begin{fullwidth}
\includegraphics[width=\linewidth]{./paramsweep.png}
\caption{\cmnt{\textbf{A} Parameter sweep caption goes here}}
\label{fig:paramsweep}
\end{fullwidth}
\end{figure}

@@

To further investigate the interplay of genetic and intrinsic mechanisms we
simulated a seminal barrel duplication experiment
\citep{shimogori_fibroblast_2005}, first by solving on a cortical domain
comprising two separate, identically shaped cortical boundaries. In this case
one distinct cortical field per thalamic projection emerged, within one of the
two boundaries only, i.e., no duplication. However, merging the two fields to
create an extended mirror-symmetric boundary shape, with an anterior-posterior
guidance field reflected at the join, gave rise to two mirror-symmetrical
barrel fields comprising 2N barrels, i.e., two cortical fields for each
thalamic projection (Fig.\,1E). Together these results suggest that by the
misexpression of Fgf8, \cite{shimogori_fibroblast_2005}
effectively created one large barrel field rather than two distinct cortical
fields.

@@

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i) at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) \cmnt{review} at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.\,\ref{eq:comp}), and iii)  \cmnt{review} the branch density of each projection is
conserved over time (Eq.\,\ref{eq:norm}).

@@

The cortical sheet was modelled as a two dimensional hexagonal lattice, which
simplifies the computation of the 2D Laplacian. Within a boundary traced
around the edge of a
rat barrel field (Fig.\,1A) we set the hex-to-hex distance
$d$ to 0.03\,mm, which resulted in a lattice containing 6515 hexes for the
simulations shown in Figs.\,1A,C \& D and 12739 hexes for the Fgf8
misexpression study shown in Fig.\,1E. Each hex contained 82 time-dependent
variables: 41 branching densities ($a_i$) and 41 connection densities ($c_i$).
The rate of change of each of the time-dependent variables (Eqs.\,\ref{eq:dc}
\& \ref{eq:da}) was computed using a fourth-order Runge-Kutta method.

@@

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife}. The
computations described in (i), (ii) and (iii) may be found in the class method
\code{RD\_James::compute\_divJ()} which calculates \code{term1}, \code{term2}
and \code{term3}, respectively.

@@

Movie corresponding to Fig.\,1C in the main paper. Simulation parameters were
$N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
$\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
projection for which the connection density is maximal, black lines delineate
boundaries, and overlaid contours show $c>0.5$. The final frame in the movie
is step 25,000 of the simulation.

>>>>

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,\ref{fig:main}A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers.  However, the
assumption of pre-arranged center-points is difficult to resolve with the
observation that axons arrive in the cortical plate as an undifferentiated
bundle, \emph{prior} to barreloid formation \citep{agmon_organized_1993}.

@@

\begin{figure}
  \begin{fullwidth}
    %\includegraphics[width=\linewidth]{./MainFig.png}
    \includegraphics[width=\linewidth]{./MainFig_superlowres.png}
    \caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
      by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
      measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
      perfect Voronoi tesselation. Right shows the initial distribution of axon
      branching density ($a$) for one thalamocortical projection, and two
      molecular guidance fields ($\rho$), \cmnt{where the domain $S$ has been
        traced from \textbf{A}}. \textbf{B} The strengths of interaction
      $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
      projections by the lengths of green and blue arrows respectively, assuming
      that similar fields aligned to the posterior-anterior and medial-lateral
      axes in the ventroposterior medial nucleus of the thalamus are sampled at
      the locations of putative barreloid centers (reconstructed from
      \citealp{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
      \cmnt{$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
      $\epsilon=1.2$} and $\delta{t}=0.0001$. Colours indicate the thalamic
      projection for which the connection density is maximal and black lines
      delineate boundaries (see Movie
      S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
      simulation approaching that obtained from barrels in \textbf{A} (dotted
      line); black squares \cmnt{give a} measure \cmnt{of} the correspondence
      between the real and
      simulated barrel shapes, $\eta$ (the sum of differences in
      area; units \cmnt{mm$^2$});
      \cmnt{grey hexagons show the measure of localization, $\omega =
        \oiint_{S} \left( \mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i\right) \mathrm{d}S$. }
      \textbf{E} \cmnt{The localization measure plotted across the domain
        shows that the structure in the connection density which has developed
        by $t=50000$ is largely absent at $t=1000$. The
        colour map
        indicates the values of  $\mathrm{max}_i(c_i)/\sum_{i=1}^{N} c_i$. Overlaid
        contours show $c_i > 0.95 \mathrm{max}_i(c_i)$.} All scale bars 1mm.}
    \label{fig:main}
  \end{fullwidth}
\end{figure}

@@

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. As such, two orthogonal linear thalamic
gradients were defined, from which 41 pairs of $\gamma$ values were sampled,
at the coordinates of 41 barreloid centers recreated from
\cite{haidarliu_size_2001} (Fig.\,\ref{fig:main}B).

A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a(\mb{x},0)$ and $c(\mb{x},0)$, a clear Voronoi-like tesselation emerged
(Fig.\,\ref{fig:main}C; see Movie S1). A reduction in the Honda delta metric (see
\citealp{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged
within \cmnt{$\approx 10000$} iterations (Fig.\,\ref{fig:main}D). A measure of the difference
in shape between real and simulated barrels revealed a strong correspondence
(see Fig.\,\ref{fig:main}D).

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./paramsweep.png}
    \caption{\cmnt{\textbf{A} Parameter sweep caption goes here}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

@@

% This figure is text width
\begin{center}
  \includegraphics[width=\linewidth]{./Fig3.png}
  \captionof{figure}{Guidance
    fields and emergent barrel pattern in a Fgf8 misexpression experiment
    (c.f. \citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
    $\rho_1$ from Fig.\,\ref{fig:main}A at the join ($\epsilon=80$).}
  \label{fig:fgf8}
\end{center}

To further investigate the interplay of genetic and intrinsic mechanisms we
simulated a seminal barrel duplication experiment
\citep{shimogori_fibroblast_2005}, first by solving on a cortical domain
comprising two separate, identically shaped cortical boundaries. In this case
one distinct cortical field per thalamic projection emerged, within one of the
two boundaries only, i.e., no duplication. However, merging the two fields to
create an extended mirror-symmetric boundary shape, with an anterior-posterior
guidance field reflected at the join, gave rise to two mirror-symmetrical
barrel fields comprising 2N barrels, i.e., two cortical fields for each
thalamic projection (Fig.\,\ref{fig:main}E). Together these results suggest that by the
misexpression of Fgf8, \cite{shimogori_fibroblast_2005}
effectively created one large barrel field rather than two distinct cortical
fields.

@@

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i) at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.\,\ref{eq:comp}), and iii) \cmnt{REVIEW: No longer true. Try plotting $a$
  vs $t$} the branch density of each projection is conserved over time.

@@

The cortical sheet was modelled as a two dimensional hexagonal lattice, which
simplifies the computation of the 2D Laplacian. Within a boundary traced
around the edge of a
rat barrel field (Fig.\,\ref{fig:main}A) we set the hex-to-hex distance
$d$ to 0.03\,mm, which resulted in a lattice containing 6515 hexes for the
simulations shown in Figs.\,\ref{fig:main}A,C \& D and 12739 hexes for the Fgf8
misexpression study shown in Fig.\,\ref{fig:fgf8}. Each hex contained 82 time-dependent
variables: 41 branching densities ($a_i$) and 41 connection densities ($c_i$).
The rate of change of each of the time-dependent variables (Eqs.\,\ref{eq:dc}
\& \ref{eq:da}) was computed using a fourth-order Runge-Kutta method.

@@

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife}. The
computations described in (i), (ii) and (iii) may be found in the class method
\code{RD\_James\cmnt{\_comp2}::compute\_divJ()} which calculates \code{term1}, \code{term2}
and \code{term3}, respectively.

@@

Movie corresponding to Fig.\,\ref{fig:main}C in the main paper. Simulation parameters were
$N=41$, $\alpha=3.6$, $\beta=16.67$, $k=3$, $D=0.5$, $\gamma\in\pm 2$,
$\epsilon=1.2$ and $\delta{t}=0.0001$. Colours indicate the thalamic
projection for which the connection density is maximal, black lines delineate
boundaries, and overlaid contours show $c>0.5$. The final frame in the movie
is step 25,000 of the simulation.
